---
header-includes:
  - \usepackage[L7x]{fontenc}
bibliography: bibliography.bib
output:
 pdf_document: 
    fig_caption: yes
    keep_tex: yes
    number_sections: no
---

```{r setup, include=FALSE}
library(plm)
library(RCurl)
library(dplyr)
library(plyr)
# library(emibase)
library(plotrix)
# library(mgcv)
library(reshape2)
library(readxl)
library(knitr)
library(rworldmap)
library(cshapes)
library(countrycode)
library(texreg)
library(ggplot2)
library(directlabels)
# library(splines)
library(MASS)
library(TTR)
library(rsdmx)
library(httr)
library(splines)
library(xtable)
# library(dynpanel)
library(devtools)
library(corrplot)
source('functions.R')
library(car)
library(tseries)
```

\vskip 20pt
\centerline{\bf \large VILNIUS UNIVERSITY}
\bigskip
\centerline{\large \textbf{FACULTY OF MATHEMATICS AND INFORMATICS}}
\vskip 120pt
\centerline{\bf \Large \textbf{BACHELOR THESIS}}
\vskip 50pt
\begin{center}
{\bf \LARGE Tourism model in the western world}

\vspace{4mm}

\end{center}
\vskip 120pt
\centerline{\Large Eligijus Bujokas}
\vskip 120pt
\centerline{\large \textbf{VILNIUS (2016.12.07)}}

\newpage

\begin{titlepage}
\centerline {\bf \large FACULTY OF MATHEMATICS AND INFORMATICS}
\centerline {\bf DEPARTMENT OF MATHEMATICAL ANALYSIS}
\vskip 120pt
\large Scientific supervisor Vaidotas Zemlys - Balevičius, PhD \underline{\hskip 95pt}
\vskip 150pt
\end{titlepage}

\newpage

\begin{center}{\large\textbf{Turizmo modelis vakarų šalyse}}\end{center}
\begin{small}
\vspace{2\baselineskip}
\begin{center}\textbf{Santrauka}\end{center}

Šio darbo tikslas yra sudaryti turizmo modelį Vakarų Europoms šalims. Naudojantis pasaulinio banko (World Bank), OECD ir Merilando Universiteto duomenų baze (the Department of Homeland Security Center of Excellence led by the Maryland University) duomenimis nustatėme, kad turistų skaičius tiesiškai nepriklauso nuo teroristinių įvykių skaičiaus, kaimyninių valstybių metinių darbo valandų skaičiaus ir nuo BVP (bendrasis vidaus produktas) vienam asmeniui. Didžiausias rastas faktorius, darantis įtaką turizmo srautui, yra šalių, iš kurių atvyksta turistai, ekonominė situacija. Tiesinės priklausomybės įrodymui (ir paneigimui) naudojome panelinius ir paprastus regresijos modelius.

\vspace{\baselineskip}
\noindent\textbf{Raktiniai žodžiai :}
 BVP, OECD, pasaulinis bankas, panelinis modelis, darbo valandos, terorizmas, turizmas
\end{small}
\vspace{\baselineskip}
\thispagestyle{empty}
\begin{center}{\large\textbf{Tourism model in western countries}}\end{center}
\begin{small}
\vspace{2\baselineskip}
\begin{center}\textbf{Abstract}\end{center}
The aim of this paper is to construct a model to help explain the number of tourist arrivals in Western European countries. By using data from the World Bank, OECD and the Department of Homeland Security Center of Excellence led by the Maryland University's database we have found that the number of tourist arrivals to a certain western country is linearly not dependant from GDP per capita, terrorist acts and the number of annual hours worked in neighbouring countries. The biggest factor to tourist arrivals was the economic state of the countries from which the tourist come from. To proove (and disproove) the linear dependance we used a panel and a simple regression model.

\vspace{\baselineskip}
\noindent\textbf{Key words :}
GDP per capita, OECD, World Bank, panel model, hours worked, terrorizm, tourism 
\end{small}
\vspace{\baselineskip}

\newpage

\tableofcontents


```{r constants, include=FALSE}
years.to.survey <- 1995:2015
path <- "data/"
# regions <- read.csv(file="input/regions.csv", stringsAsFactors = F)
createdir("plots")
# createdir("input")
createdir("data")
createdir("output")
# devtools::install_url("http://cran.r-project.org/src/contrib/rprojroot_1.2.tar.gz") ## MAY RESOLVE TROUBLESHOOTS
knitr::opts_chunk$set(fig.pos = 'h')
```

```{r data download and reading it, include=FALSE, echo=F}
# we will save this data to the 'data' folder
download.terror(years.to.survey, "data/") 
dt <- read.terror("data/terror/")
dt <- arrange(dt, Date)
na.index <- which(apply(dt, 1, function(x){ all(is.na(x)) }))
if(length(na.index)!=0) dt <- dt[-na.index, ]
write.csv(aggregate.terror.by.country(dt), file="input/terror raw data.csv", na="", row.names=F)

## Tourism data
# 
# dt.tour <- download.tourism("data/")
# write.csv(dt.tour, file="input/total arrivals.csv",  na="", row.names=F)

## Economic data

dt.eco <- download.economy("data/")
dt.eco <- make.economy.great.again(dt.eco)
write.csv(dt.eco, file="input/gdp.csv", na="", row.names=F)

## OECD data on hours worked

dt.oecd <- get_dataset("ANHRS", path="data/")
decode  <- get_decoder("data/")  
write.csv(dt.oecd, file="input/hours.csv", na="", row.names=F)

```

```{r data input, include=FALSE, echo=F}

decoder  <- read.csv("input/decoder.csv", stringsAsFactors = F)
struct <- read.csv("input/tourism structure.csv", stringsAsFactors = F) %>% get.cols
flows  <- read.csv("input/receipts.csv", stringsAsFactors = F) %>% get.cols
terror <- read.csv("input/terror raw data.csv", stringsAsFactors = F)
tax    <- read.csv("input/tax.csv", stringsAsFactors = F) %>% filter(MEASURE=="PC_TOT_TAX") 
gdp    <- read.csv("input/gdp.csv", stringsAsFactors = F) %>% rename(c("CountryName" = "Country"))
hours  <- read.csv("input/hours.csv", stringsAsFactors = F) %>% format.hours(decoder)
total.arrivals <- read.csv("input/total arrivals.csv", stringsAsFactors = F)

```

```{r formating input for the big model, include=FALSE, echo=F}

## Merging everything to one data frame

master.data <- merge(total.arrivals, terror)
gdp.m <- gdp[, c("Country", "Date", "value")] %>% rename(c("value"="gdp"))
master.data <- merge(master.data, gdp.m)

tax <- rename(tax, c("LOCATION" = "ï..LOCATION"))
tax <- tax[, c("ï..LOCATION", "TIME", "INDICATOR",  "Value")]
tax <- rename(tax, c("ï..LOCATION" = "Country", "INDICATOR" = "Indicator", "TIME" = "Date"))
  
tax <- ddply(tax, ~Country, function(xframe){
    
  xframe <<- xframe
  real.name <-  decoder[decoder$CountryCode==xframe$Country[1], "CountryName"]
  if(length(real.name)==0) real.name <- xframe$Country[1]
  xframe$Country <- real.name
  return(xframe)
    
})
  
tax[tax$Country=="Korea, Rep.", "Country"] <- "South Korea"  
tax[tax$Country=="Korea", "Country"] <- "South Korea" 
  
tax.m <- tax[, c("Country", "Date", "Value")] %>% rename(c("Value"="tax"))
master.data <- merge(master.data, tax.m)
hours.m <- hours[, c("Country", "Date", "Value")] %>% rename(c("Value"="hours.worked"))
master.data <- merge(master.data, hours.m)
master.data <- master.data[complete.cases(master.data), ]

master.data <- ddply(master.data, ~Country, function(xframe){
  
  xframe <<- xframe
  if(nrow(xframe)!=length(master.data$Date %>% unique())){
    
    xframe <- NULL
    
  }
  
  return(xframe)
  
})

countries <- c("France", "United Kingdom", "United States", "Canada", "Germany", "Spain", "Portugal", "Netherlands", "Austria", "Italy", "Belgium")

master.data <- master.data[master.data$Country %in% countries, ]

pdata <-  plm.data(master.data, index = c("Country", "Date")) 
log.vars <- c("Total.Arrivals", "gdp", "hours.worked")
pdata[, log.vars] <- apply(pdata[, log.vars], c(1, 2), log)

pdata <- ddply(pdata, ~Country, function(xframe){
  
  xframe <<- xframe
  xframe[, 3:dim(xframe)[2]] <- apply(xframe[, 3:dim(xframe)[2]], 2, function(x){
  
  return(c(NA, diff(x)))
  
  })
  
  return(xframe)
  
})

pdata <- pdata[complete.cases(pdata), ]
```
\newpage

\section{Introduction}

Wikipedia gives the following definition regarding what is tourism [see @tourism.definition]: "Tourism is travel for pleasure; also the theory and practice of touring, the business of attracting, accommodating, and entertaining tourists, and the business of operating tours". According to UNWTO (United Nations World Tourism Organization)[see @UNWTO]:

$\bullet$ 1 in 11 jobs worldwide are related to the tourism industry.

$\bullet$ 10 percent of total world GDP is produced by tourism.

$\bullet$ Tourism made up for about 7 percent of world's exports: 1.5 trillion US dollars.

$\bullet$ In 2015 there were 1186 million tourists worldwide and this number is expected to grow up to 1.8 billion by 2030.

$\bullet$ In 2015 the total receipts by tourists was about 1.26 billion US dollars.

From these figures it is clear to see that every country can benefit from an increase in tourism flows. Tourism continues to demonstrate its key role in generating economic activity, employment and export revenues in the OECD area, where it directly contributes, on average, 4.1 \% of GDP, 5.9 \% of employment and 21.3 \% of service exports. Tourism offers strong potential to support job rich growth and at around 80 percent, tourism exports also generate higher than average domestic value added. International tourist arrivals surpassed 1.1 billion in 2014 (World Tourism Organization), following a resurgence in arrivals to OECD countries (6.4 \%), which increased at a faster rate than the global average (4.2 \%) (as of 2016). A large part of countries that are regarded as the "Western World" are members of the OECD organization. In this paper we will hold that the western countries are France, United Kingdom, USA, Canada, Germany, Spain, Portugal, Netherlands, Austria, Italy and Belgium.

In this paper we want to find which economic or social variables significantly influence the tourism flows to help governments in deciding which policies to adapt in order to grow the tourism sector.

\newpage

\section{Countries}

The countries which will be analyzed are all members of the OECD (\textbf{Organisation for Economic Co-operation and Development}) [see @OECD].  

```{r map of OECD countries, echo=F, fig.width=8, fig.height=8, fig.cap= "Map of the countries in this analysis", echo=F, warning=F, include=T}

OECD.cn <- data.frame(country = unique(master.data$Country), 
                      OECD = rep.int("Western countries", 
                                     length(unique(master.data$Country))))
malMap <- joinCountryData2Map(OECD.cn, joinCode = "NAME",
  nameJoinColumn = "country", verbose = F, suggestForFailedCodes = F)

mapCountryData(malMap, nameColumnToPlot="OECD", catMethod = "categorical",
  missingCountryCol = gray(.8), oceanCol = 'cyan', mapTitle = "Political map of the world")


```

\newpage

\subsection{Total number of arrivals}

To have a better picture of the tourist flows to the countries in this analysis we can visualize the data.

```{r arrivals visualization, echo=F, fig.cap="Total number of tourist arrivals", fig.height=10, fig.width=11, fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "Total.Arrivals"), ]

dt.a$Date <- as.numeric(dt.a$Date)
dt.a$Total.Arrivals <- dt.a$Total.Arrivals/1000

p <- ggplot(dt.a, aes(x=Date, y=Total.Arrivals, group=Country, colour=Country)) + geom_line(size=1.2) + ylab("Total Arrivals, thousands") + theme_bw()  + theme(legend.position="none") + scale_x_continuous(expand = c(0, 3)) + geom_dl(aes(label = Country), method = list(dl.combine("last.bumpup"), cex = 1)) 

plot(p)

```

To see the overal trend from all of these countries is quite difficult so we will plot the sum of tourist arrivals in each year.

```{r arrival trend, echo=F, fig.cap="Trend of tourists arrivals", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "Total.Arrivals"), ]

dt.a <- ddply(dt.a, ~Date, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country='All', Date=xframe$Date[1], Total.Arrivals=sum(xframe$Total.Arrivals, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a$Total.Arrivals <- dt.a$Total.Arrivals/1000
dt.a$Date <- as.numeric(dt.a$Date)

p <- ggplot(dt.a, aes(x=Date, y=Total.Arrivals, group=Country, colour=Country, label=Country)) + geom_line(size=1.2) + ylab("Total Arrivals, thousands") + theme_bw() + scale_x_continuous(expand = c(0, 4)) + geom_dl(aes(label = Country), method = list(dl.combine("first.polygons"), cex = 1)) + geom_dl(aes(label = Country), method = list(dl.combine("last.polygons"), cex = 1)) + stat_smooth(se = FALSE, method = "lm", formula = y ~ ns(x,1),aes( colour="Linear trend"), linetype="dashed")  + scale_colour_manual(values = c("coral3", "darkblue"),
                       guide = guide_legend(override.aes = list(
                         linetype = c("solid", "dashed"),
                         shape = c(16, 16)), title="Legend"))

plot(p)

```

From figure 3 we can clearly see that the overal sum of tourist arrivals through the years is increasing rapidly. In this paper we will try to model these arrival values and identify the biggest factors of this process. 

\newpage
\section{Economic variable selection}

The aim of this paper is to create a model to help explain the total number of arrivals of tourists to a given OECD country. We want to find the function \textbf{f} in the equation:

\[
log(A)_{it}  = \textbf{f}(\Delta log(GDP)_{it}, \Delta terror_{it}, \Delta log(hours)_{it}, \Delta tax_{it})
\]

Here 

$A_{it}$ - total number of arrivals to a country i at time period t, thousands

$GDP_{it}$ - GDP per capita at a country i at time period t, thousands 

$terror_{it}$ - total number of terrorist attacks at country i at time period t

$hours_{it}$ - average numbers of hours worked annualy in country i at time period t

$tax_{it}$ - percentage of taxes on goods and services.

<!-- \subsection{Motivation behind variable selection} -->

\newpage

\subsection{Gross domestic product per capita}

The variable $GDP$ was chosen to reflect the economic situation and the standart of living in a given country. GDP per capita is calculated by taking the  total output of a country and dividing it by the number of people in the country. The division by the number of people helps us to compare economies between countries. 

```{r OECD GDP per capita, echo=F, fig.cap="GDP per capita of OECD", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "gdp")]

dt.a <- ddply(dt.a, ~Date, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country='All', Date=xframe$Date[1], gdp=mean(xframe$gdp, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a <- rename(dt.a, c("gdp" = "value"))
dt.a$Date <- as.numeric(dt.a$Date)

p <- ggplot(dt.a, aes(x=Date, y=value, group=Country, colour=Country, label=Country)) + geom_line(size=1.2) + ylab("GDP per capita, USD") + theme_bw() + scale_x_continuous(expand = c(0, 4)) + geom_dl(aes(label = Country), method = list(dl.combine("first.polygons"), cex = 1)) + geom_dl(aes(label = Country), method = list(dl.combine("last.polygons"), cex = 1))  + stat_smooth(se = FALSE, method = "lm", formula = y ~ x,aes( colour="Linear trend"), linetype="dashed") +
scale_colour_manual(values = c("darkblue", "coral3"),
                       guide = guide_legend(override.aes = list(
                         linetype = c("dashed", "solid"),
                         shape = c(16, 16)), title="Legend")) 

plot(p)

```

\newpage

\subsection{Number of terrorist attacks}

In recent years there has been an increase in terrorist attacks in the western world: France, Germany, Turkey and others have suffered attacks, especially from the group called ISIS. It is quite straightforward to think that in light of these attacks tourists will be trying to avoid countries with a higher chance of a terrorist attack. We will want to know whether terrorist attacks is a significant variable to tourist arrivals and how much of an impact it has. 

```{r visualizing terror, echo=F, fig.cap="Terrorism in OECD countries", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "Terror.attacks")]

dt.a <- ddply(dt.a, ~Country, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country=xframe$Country, Date="Total", Terror.attacks=sum(xframe$Terror.attacks, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a <- t(dt.a[, c("Country", "Terror.attacks")])
# dt.a[2, ] <- as.numeric(dt.a[2, ])
nn <- dt.a[1, ]
dt.a <- dt.a[-1, ] %>% as.data.frame() %>% t()
dt.a <- apply(dt.a, c(1,2), function(x) {as.numeric(x)})
names(dt.a) <- nn
dt.a <- sort(dt.a)
par(mai=c(1,2,1,1))
barplot(dt.a , col='deepskyblue3',  horiz=T , las=1, cex.names = 0.7, main="Total number of attacks, 1996 - 2014")

```

\newpage

\subsection{Average annual working hours}

Most countries in the developed world have seen average hours worked decrease significantly. This is due because of the effectiveness of new tools and technologies, overall population growth and other factors.

```{r visualizing working hours, echo=F, fig.cap="Hours worked in OECD countries", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "hours.worked")]

dt.a <- ddply(dt.a, ~Date, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country='All', Date=xframe$Date[1], hours.worked=mean(xframe$hours.worked, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a <- rename(dt.a, c("hours.worked" = "value"))
dt.a$Date <- as.numeric(dt.a$Date)

p <- ggplot(dt.a, aes(x=Date, y=value, group=Country, colour=Country, label=Country)) + geom_line(size=1.2) + ylab("Average hours worked annualy") + theme_bw() + scale_x_continuous(expand = c(0, 4)) + geom_dl(aes(label = Country), method = list(dl.combine("first.polygons"), cex = 1)) + geom_dl(aes(label = Country), method = list(dl.combine("last.polygons"), cex = 1)) + stat_smooth(se = FALSE, method = "lm", formula = y ~ x,aes( colour="Linear trend"), linetype="dashed") +
scale_colour_manual(values = c("darkblue", "coral3"),
                       guide = guide_legend(override.aes = list(
                         linetype = c("dashed", "solid"),
                         shape = c(16, 16)), title="Legend")) 

plot(p)

```

As we can see from figure 6 the overal downwards trend of the working annaul working hours' average is true to the countries which we will analyze.

\newpage

\section{Modelling}

\subsection{Panel data structure}

Panel data provides information on individual behaviour, both across individuals and over time - the data has cross-sectional and time-series dimensions. Panel data includes N individuals over T regular time periods. We can denote a panel data model similarly like a typical bivariate reggression model [see @econometric.methods, pp. 390]:

\[
\tag{1}
  y_{it} = \alpha_{i} + \beta X_{it} + u_{it}
\]

The difference from a typical reggression model is that there are the $i$ and the $t$ terms, where $i$ denotes the individual and $t$ denotes the time period. The construction of a panel model starts with appropriately 'stacking' the data set. Assuming that y is the dependant variable and X is the matrix of the explanatory variables we need to have the following structure [see @econometric.methods, pp. 388]: 

\[
  Y = y_{i} = \begin{bmatrix}
         y_{1}    \\
         y_{2}    \\[0.3em]
         \vdots   \\[0.3em]
         y_{N}     
  \end{bmatrix} = \begin{bmatrix}
         e    \\
         0    \\[0.3em]
         \vdots   \\[0.3em]
         0     
  \end{bmatrix}  \alpha_{1} + \begin{bmatrix}
       0    \\
       e    \\[0.3em]
       \vdots   \\[0.3em]
       0    
\end{bmatrix}  \alpha_{2} + \cdots + \begin{bmatrix}
       0    \\
       0    \\[0.3em]
       \vdots   \\[0.3em]
       e     \\[0.3em]
\end{bmatrix}  \alpha_{N} + \begin{bmatrix}
       x_{1}    \\[0.3em]
       x_{2}    \\[0.3em]
       \vdots   \\[0.3em]
       x_{n}     \\[0.3em]
\end{bmatrix} \beta +   \begin{bmatrix}
       u_{1}    \\[0.3em]
       u_{2}    \\[0.3em]
       \vdots   \\[0.3em]
       u_{n}     \\[0.3em]
\end{bmatrix}
\]

Here

\[
y_{i} = \begin{bmatrix}
       y_{i1}    \\
       y_{i2}    \\[0.3em]
       \vdots   \\[0.3em]
       y_{iT}     \\[0.3em]
\end{bmatrix} x_{i} = \begin{bmatrix}
       x_{1i1} & x_{2i1} & \cdots & x_{Ki1}    \\
       x_{1i2} & x_{2i2} & \cdots & x_{Ki2}   \\[0.3em]
       \vdots & \vdots & \cdots & \vdots   \\[0.3em]
      x_{1iT} & x_{2iT} & \cdots & x_{KiT}      \\[0.3em]
\end{bmatrix}
\]

For example, the 'head' and the 'tail' of the raw data which will be analysed looks like this:

```{r visualising data table, echo=F}

kable(head(pdata), format = "latex", row.names = F)
kable(tail(pdata), format = "latex", row.names = F)

```


\newpage

\subsection{Unit roots}

Before modelling we need to check whether there are unit roots in our data set. We will use the \textit{purtest()} function to check for this. This function implements the Maddala and Wu test.

```{r adf, echo=T, warning=F, comment=NA}

Arrivals <- as.data.frame(split(pdata$Total.Arrivals, pdata$Country))
purtest(Arrivals, pmax = 4, exo = "intercept", test = "madwu")

GDP <- as.data.frame(split(pdata$gdp, pdata$Country))
purtest(GDP, pmax = 4, exo = "intercept", test = "madwu")

Hours <- as.data.frame(split(pdata$hours.worked, pdata$Country))
purtest(Hours, pmax = 4, exo = "intercept", test = "madwu")

Attacks <- as.data.frame(split(pdata$Terror.attacks, pdata$Country))
purtest(Attacks, pmax = 4, exo = "intercept", test = "madwu")

Tax <- as.data.frame(split(pdata$tax, pdata$Country))
purtest(Tax, pmax = 4, exo = "intercept", test = "madwu")

```

The alternative hypothesis for the test is that the data does not have a unit root. According to the tests no variable has a unit root. 

\newpage

\subsection{Pooling estimator}

The first model which we will be examining will be the simple OLS model which ignores the panel data structure. The equation is then:

\[
y = \alpha + X \beta + \epsilon
\]

And the estimator is derived solving the equation:

\[
\widehat{\beta} = \left( X X^{T} \right)^{-1}X^{T}y
\]

```{r pooling panel model, include=FALSE, echo=F, warning=F}

model.ols <- lm(formula = Total.Arrivals ~ Terror.attacks + hours.worked + tax + gdp, data=pdata)
# summary(model.ols)
```

```{r xtable, results="asis", echo=F, warning=F}
tab <- xtable(summary(model.ols)$coef, digits=c(3, 3, 3, 3, 3), caption = "OLS model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```

The problem with this model is that while giving consistent estimators, the standart errors of those estimators will be understated. Also, OLS is not efficient compared to the generalized least-squares procedure [see @econometric.methods, pp. 391].

\newpage

\subsection{Fixed effects model}

In this section we will be creating a fixed effects panel data model [see @econometric.methods, pp. 395 - 399]. Recall equation (1). The $\alpha_{i}$ is called the \textit{individual effect}.
If $\alpha_{i}$ does correlate with $X_{it}$ then the panel model can be reffered to as a fixed effects model. 

\subsubsection{Model assumptions}

Let as assume that we have n individuals and T observations for each individual. The fixed effects model has the following structure:

$$y_{it} = X_{it} \beta + \epsilon_{it} $$
$$\epsilon_{it} = \alpha_{i} + \eta_{it} $$

The assumptions for the errors are:

$$\mathbf{E}[\eta] = 0 ; E[\eta\eta^{\prime}] = \sigma^{2}_{\eta}I_{nT}$$

$$\mathbf{E}[\alpha_{i}\alpha_{j}] = 0, \forall i\neq j;  E[\alpha\alpha^{\prime}] = \sigma^{2}_{\alpha}$$

$$\mathbf{E}[\alpha_{i}\eta_{j}] = 0, ;  E[\alpha] = 0$$

All of the above expectations are conditional on $X$. In the fixed effects model case the $\alpha_{i}$ term is treated as an unknown parameter and has to be estimated. 

\subsubsection{Obtaining estimates}

To obtain the $\widehat{\beta_{FE}}$ we first "demean" the data. $\forall i$:

$$y_{it} - \overline{y_{i}} = \left(X_{it} -  \overline{X_{i}}\right) \beta + \left( \alpha_{i}  - \overline{\alpha_{i}}\right) + \left( \epsilon_{i}  - \overline{\epsilon_{i}}\right)$$

Now since $\overline{\alpha_{i}}$ is just $\alpha_{i}$ then the above equations simplifies to:

$$\dot{y} := y_{it} - \overline{y_{i}} = \left(X_{it} -  \overline{X_{i}}\right) \beta + \left( \epsilon_{i}  - \overline{\epsilon_{i}}\right) =: \dot{X} \beta_{FE} + \dot{\epsilon}$$

Now to obtain the estimate $\widehat{\beta_{FE}}$ we simply run an OLS procedure to the equation above. The estimator $\widehat{\beta_{FE}}$ is often called the within estimator. Any procedure that eliminates the term $\alpha_{i}$ can be called a $\textit{within transformation}$. To obtain the $alpha_{i}$ for every $i$ we simply match the intercept coordinates that is estimated by running the reggresion for the demeaned data with the way the data was stacked. 

\newpage

\subsubsection{Results}

```{r fixed effects model, include=FALSE, echo=F}

model.fixed <- plm(formula = Total.Arrivals ~ Terror.attacks + hours.worked + tax + gdp, data=pdata, model="within")
# summary(model.fixed)
```

```{r, results='asis', echo=F, warning=F}

tab <- xtable(summary(model.fixed)$coef, digits=c(3, 3, 3, 3, 3), caption = "Fixed effects model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)

```

As we can see from table 2 the coefficients have a simmilar signs to them as in the OLS case. The only significant variable is the average hours worked variable. One percent increase would lead to an increase of about the same amount in tourist arrivals.

After dropping the insignificant variables we have:


```{r final models, include=FALSE, echo=F}

model.fixed <- plm(formula = Total.Arrivals ~ hours.worked, data=pdata, model="within")
model.ols <- lm(formula = Total.Arrivals ~ hours.worked, data=pdata)
# summary(model.fixed)
```

```{r, results='asis', echo=F, warning=F}

tab <- xtable(summary(model.fixed)$coef, digits=c(3, 3, 3, 3, 3), caption = "Fixed effects model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)

tab <- xtable(summary(model.ols)$coef, digits=c(3, 3, 3, 3, 3), caption = "OLS model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```

Applying the F test for individual effects gives the result:

```{r pFtest, echo=T, comment=NA}
  pFtest(model.fixed, model.ols)
```

According to these results the fixed effects model is not significantly better in explaining the number of arrivals compared to the OLS model.

```{r estimating ols and fixed model fits, echo=F, comment=NA}
  
  pdata <- ddply(pdata, ~Country + Date, function(xframe){
    
    xframe <<- xframe
    
    
    ## fixed effects
    
    intercept <- fixef(model.fixed)[xframe$Country[1]] %>% as.numeric()
    X <- xframe[, c('hours.worked')]
    fit <- model.fixed$coefficients[1] * X[1] + intercept
    fit <- as.numeric(fit) # this is the fit of the logged number of arrivals
    xframe$fc.fixed.effect <- fit
    
    ## OLS
    
    X <- xframe[, c('hours.worked')] 
    fit <- model.ols$coefficients[2] * X[1]
    fit <- fit + model.ols$coefficients[1] 
    xframe$fc.OLS <- as.numeric(fit)
    
    return(xframe)
  })

```


```{r visualizing the differences between fits, echo=F, fig.cap="OLS and fixed effects model comparison", fig.height=7, fig.width=8,  fig.pos="H"}
  
fdt <- pdata[, c("Country", "Date", "Total.Arrivals", "fc.fixed.effect", 'fc.OLS')]
# fdt$Total.Arrivals  <- exp(fdt$Total.Arrivals)/1000
# fdt$fc.fixed.effect <- exp(fdt$fc.fixed.effect)/1000
# fdt$fc.OLS          <- exp(fdt$fc.OLS)/1000
fdt$Date <- fdt$Date %>% as.character() %>% as.numeric()

## we will plot 4 countries in order to have a clear picture
cn <- c( "United States",  "Austria", "France", "Spain")
points <- data.frame()

par(mfrow=c(2,2))
for(cc in cn){
  
  tmp <- fdt[fdt$Country==cc, ]
  points <- tmp[, c("Total.Arrivals", "fc.fixed.effect", 'fc.OLS')]
  
grid.frame(x=as.numeric(tmp$Date), y=points)
  matplot(x=as.numeric(tmp$Date), y=points, lwd=1, lty=1, cex=1.25, pch=20, xlab="Time", ylab="Arrivals, thousands", add = T, type="o",
        col=c('salmon1', "firebrick1", 'cornflowerblue'))
  legendary2(c("Original", "Fixed effect fit", "OLS fit"), col=c('salmon1', "firebrick1", 'cornflowerblue'))
  mtext(cc, col="blueviolet", line=2, cex=1.25, adj = 0)
}

```

\newpage

\subsection{Statistics for accuracy}

We can measure the goodness of fit using several statistics. We will use the the \textit{mean absolute error} (MAE), \textit{mean squared error} (MSE)
and the \textit{mean absolute percentage error} (MAPE) [see @accuracy]. These statistics are defined by:

\[
MSE = \dfrac{1}{n}  \sum_{i=1}^{n}\left( y - \widehat{y} \right)^{2}
\]

\[
MAE = \dfrac{1}{n}  \sum_{i=1}^{n}\left| y - \widehat{y} \right|
\]

\[
MAPE = \dfrac{100}{n}  \sum_{i=1}^{n}\left| \dfrac{y - \widehat{y}}{y} \right| 
\]

Here

$\widehat{y}$ - fitted value

$y$ - original value


$n$ - number of observations


\begin{center}

```{r forecast accuracy, echo=F, results= "asis", fig.pos="H"}
  
fdt <- pdata[, c("Country", "Date", "Total.Arrivals", "fc.fixed.effect", 'fc.OLS')]
# fdt$Total.Arrivals  <- exp(fdt$Total.Arrivals)
# fdt$fc.fixed.effect <- exp(fdt$fc.fixed.effect)
# fdt$fc.OLS          <- exp(fdt$fc.OLS)

accuracy.table <- matrix(ncol=2, nrow=3) %>% as.data.frame()
colnames(accuracy.table) <- c("OLS", "Fixed effects")
rownames(accuracy.table) <- c("MSE", "MAPE", "MAE")

## MSE:

accuracy.table[1, 1] <- MSE(fdt$Total.Arrivals, fdt$fc.OLS)
accuracy.table[1, 2] <- MSE(fdt$Total.Arrivals, fdt$fc.fixed.effect)

## MAPE

accuracy.table[2, 1] <- MAPE(fdt$Total.Arrivals, fdt$fc.OLS)
accuracy.table[2, 2] <- MAPE(fdt$Total.Arrivals, fdt$fc.fixed.effect)

## MAE

accuracy.table[3, 1] <- MAE(fdt$Total.Arrivals, fdt$fc.OLS)
accuracy.table[3, 2] <- MAE(fdt$Total.Arrivals, fdt$fc.fixed.effect)

# options("scipen" = 5)

accuracy.table <- apply(accuracy.table, c(1, 2), function(x) {
  # x <- format(x,scientific=FALSE)
  x <- as.numeric(x)
  x <- round(x, 3)
  return(x)
  })

tab <- xtable(accuracy.table[c(1,3), ], digits=c(3, 3, 3), caption = "Accuracy statistics results")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```

\end{center}

As we can see from table 5 and the statistics for accuracy table the fixed effects models' predictions are closer to the original values. Note that we have dropped the MAPE statistic because we are measuring percentage changes and the statistic would give misleading results.

\newpage

\subsection{Random effects model}

There are two main types of panel models: the fixed effects model and the random effects models [see @econometric.methods, pp. 391 - 393]. The main difference between them is the assumption of $\alpha_{i}$ regarding the regressor matrix $X_{it}$. In the random effects model case it is assumed that $\alpha_{i}$ and $X_{it}$ do not correlate.

```{r random effects model, include=FALSE, echo=F}

model.random <- plm(formula = Total.Arrivals ~ Terror.attacks + gdp + tax + hours.worked , data=pdata, model="random")
# summary(model.random)
```
```{r, results='asis', echo=F, warning=F}
  # texreg(model.random, caption = "Random effects model", custom.coef.names = c("intercept", "attacks", "log GDP", "log Hours"), digits = 3)

tab <- xtable(summary(model.random)$coef, digits=c(3, 3, 3, 3, 3), caption = "Random effects model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```

Comparing tables 2 and 6 we can see that the signs near the coefficients are the same and the actual values of the estimators are almost even. The random effects model has a little higher R statistic. The main difference here is that there is the intercept term and we lose some "uniqueness" to the countries which was present with the fixed effects model. One measure to determine which model is better for the given data is the Hausman test. Just like in the previous cases, the only significant variable is the hours worked variable. The final random effects model:


```{r random effects model 2, include=FALSE, echo=F}

model.random <- plm(formula = Total.Arrivals ~ hours.worked, data=pdata, model="random")

```
```{r, results='asis', echo=F, warning=F}

tab <- xtable(summary(model.random)$coef, digits=c(3, 3, 3, 3, 3), caption = "Random effects model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```


```{r pHtest, echo=T, comment=NA}
  phtest(model.random, model.fixed)
```

As we can see from the test result we cannot reject the null hypothesis therefore we conclude that the random effects coefficients are more efficient and are consistent. Therefore the test implies that we sould use the random effects model.

\newpage

\subsection{Residuals}

As we saw in the previous section the Hausman test for panel data indicated that we should use the random effects model. The test is only one of the criteria for deciding which model to use. Two critical assumptions of any linear model, including linear fixed-effects panel models and the random effects models, are homoscedastic and normally distributed errors.

```{r residuals of models, echo=F, fig.cap="Residuals of models", fig.height=7, fig.width=8,  fig.pos="H"}
  
par(mfrow=c(1,2))

res.random <- residuals(model.random)
res.fixed  <- residuals(model.fixed)
  
grid.frame(x=seq(from=1, length.out = length(res.random)), y=as.numeric(res.random))
matplot(x=seq(from=1, length.out = length(res.random)), y=as.numeric(res.random), lwd=1, lty=1, cex=1.25, pch=20, ylab="Residual value", xlab="Index", type="o", add=T, col=c('salmon1'))
mtext("Random effects model residuals", col="salmon1", line=1, cex=1.25, adj = 0)

grid.frame(x=seq(from=1, length.out = length(res.random)), y=as.numeric(res.random))
matplot(x=seq(from=1, length.out = length(res.fixed)), y=as.numeric(res.fixed), lwd=1, lty=1, cex=1.25, pch=20, ylab="Residual value", xlab="Index", type="o", col=c('cornflowerblue'), add=T)
mtext("Fixed effects model residuals", col="cornflowerblue", line=1, cex=1.25, adj = 0)

```

As we can see in figure 8 the residuals visually look very similar. 

```{r distribution of residuals, echo=F, fig.cap="Distribution of the residuals", fig.height=7, fig.width=8,  fig.pos="H"}
  
par(mfrow=c(2,2))

res.random <- residuals(model.random)
res.fixed  <- residuals(model.fixed)
  
hist(res.random, col = "cornsilk3", xlab = "Residuals of the random model", main="Density of the residuals")
hist(res.fixed, col = "cornsilk3", xlab = "Residuals of the fixed effects model", main="Density of the residuals")

qqnorm(res.random, ylab='Residuals', main = "Q - Q plot for random effects model", col="cornflowerblue")
grid()
qqline(res.random)

qqnorm(res.fixed, ylab='Residuals', main = "Q - Q plot for fixed - effects model", col="cornflowerblue")
grid()
qqline(res.fixed)

```

Figure 9 shows the histogram and the quantile to quantile plot of the residuals. In both cases, the residuals appear to follow a bell - shaped curve which implies normality. Altough the q q plots show that the residuals have heavy tails. This can be expected because real life data very often is not 'clean' enough to give normally distributed errors. 

Both the random effects models and fixed effects models are not satisfactory because of the residuals, low R - square and the fitted values. Additionally, the random effect models' underlying assumption of $\alpha_{i}$ not beeing correlated with the explanatory variables is often not true. In the next section we will switch to a dynamic panel model.

\newpage

\subsection{Dynamic panel data model}


In this section we will be creating a dynamic panel model (we will include lags of the variables). First of all, the estimation of the coefficient $\delta$ in the equation:

$$ y_{it} = \delta y_{it - 1} + \beta X_{it} + \alpha_{i} + \epsilon_{it} $$

is not as straightforward as just adding another collumn to the regressor matrix with the lagged variable $y_{it}$. This is because the OLS estimate will be biased. This is because the $y_{it}$ is a function $\epsilon_{it}$ and thus so is $y_{it-1}$ giving us the multicolinearity problem. 

Consider the simple model:

$$y_{it} = \delta y_{it - 1}+ \alpha_{i} + \epsilon_{it}$$

Taking the first differences we get rid of the individual effecct $\alpha_{i}$:

$$\Delta y_{it} = y_{it} - y_{it - 1} = \delta y_{it-1} + \alpha_{i} + \epsilon_{it} - \delta y_{it - 2} - \alpha_{i} - \epsilon_{it-1} = \delta (y_{it-1} - y_{it-2}) + \epsilon_{it} - \epsilon_{it-1} $$
The classic method to deal with multicolinearity is to find an instrumental variable that correlates with the endogenous variable and not with the residuals. Starting from time period t = 3:

$$\Delta y_{i3} = y_{i3} - y_{i2} = \delta (y_{i2} - y_{i1}) + (\epsilon_{i3} - \epsilon_{it-2}) $$
Here we can see that the variable $y_{i1}$ does not correlate with $(\epsilon_{i3} - \epsilon_{i2})$. 

When t = 4:

$$\Delta y_{i4} = y_{i4} - y_{i3} = \delta (y_{i3} - y_{i2}) + (\epsilon_{i4} - \epsilon_{i3}) $$
Both the $y_{i1}$ and ${y_{i2}}$ are good options. Going forward up to time period T we have the instrument vector:

$$(y_{i1}, y_{i2}, ..., y_{iT- 2})$$. If there are strictly exogenous variables in our system $x_{it}$ that correlate with the endogenous variable and not with the residuals they should also be included in the weight matrix diagonals (for a detailed proof [see @dynamic.panel]):


$$W_{i} = \begin{pmatrix}
    [y_{i1}, x_{i1}, x_{i2}]       & 0 & \dots & 0 \\
    0       &  [y_{i1}, y_{i2} x_{i1}, x_{i2}, x_{i3}] & \dots & 0 \\
    0      & 0  & \dots & [y_{i1}, y_{i2}, ..., y_{iT-2}, x_{i1}, x_{i2}, x_{i3}, ..., x_{iT-1}]
\end{pmatrix} $$

Then the consistent Arriano-Bond estimator is given by:

$$\widehat{\delta} =  \left[ (\Delta y_{t - 1})^{t} W (\widehat{V}_{W})^{-1} W^{t} \Delta y_{t - 1}\right]^{-1} x \left[ (\Delta y_{t - 1})^{t} W (\widehat{V}_{W})^{-1} W^{t} \Delta y_{t - 1}\right] $$
$$ \widehat{V}_{W} = \sum_{i=1}^{n} W_{i}^{t} \Delta \epsilon_{i} \Delta \epsilon_{i}^{t}  W_{i}$$

```{r dynamic model, include=FALSE, echo=F}

model.dynamic <- suppressWarnings( pgmm(Total.Arrivals ~ Terror.attacks + hours.worked + tax + gdp,
                       lag.form=list(1, 0, 0, 0, 0), gmm.inst = ~ Terror.attacks + hours.worked + tax + gdp, 
                       lag.gmm = list(c(2,99)),
                       effect="individual",
                       data=pdata))
```
```{r, results='asis', echo=F, warning=F, fig.pos="H"}
  # texreg(model.train, caption = "Fixed effects model with training set", custom.coef.names = c("attacks", "log GDP", "log Hours"), digits = 3)

tab <- xtable(summary(model.dynamic)$coef, digits=c(3, 3, 3, 3, 3), caption = "Dynamic panel data model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```

After trying various laggs and taking into account the economic logic behind the model we get the results in table 8. We see that we get a very similar result as in the static case.

\newpage

\subsection{Dynamic models' residuals}

```{r residuals of dynamic models, echo=F, fig.cap="Residuals of models", fig.height=8, fig.width=7,  fig.pos="H"}
  
par(mfrow=c(3, 1))

res.dynamic <- model.dynamic$residuals %>% unlist
  
grid.frame(x=seq(from=1, length.out = length(res.dynamic)), y=as.numeric(res.dynamic))
matplot(x=seq(from=1, length.out = length(res.dynamic)), y=as.numeric(res.dynamic), lwd=1, lty=1, cex=1.25, pch=20, ylab="Residual value", xlab="Index", type="o", add=T, col=c('salmon1'))
mtext("Dynamic model residuals", col="salmon1", line=1, cex=1.25, adj = 0)

hist(res.dynamic, col = "cornsilk3", xlab = "Residuals of the dynamic model", main="Density of the residuals")

qqnorm(res.fixed, ylab='Residuals', main = "Q - Q plot for dynamic model", col="cornflowerblue")
grid()
qqline(res.fixed)

```
As we can see in figure 17 the residuals resemble a simmilar pattern as in the case with static panel models. Altough the quantile to quantile plot shows "lighter" tails than in the previous cases. 

\subsection{Conclusion of panel data models' analysis}

Based on the data and using various panel models we can conlude that the only significant variable to tourist arrivals' changes is changes to the average hours worked in a given country. On average, a one percent increase in hours worked would lead to an increase of 1 percent of tourist arrivals. This can be explained that more hours worked would lead to more bussiness and hence an increase in tourist attractions in a given country. 

Another important note is that we found that changes in terrorist attacks has so far made no influence to tourist flows in the western world. This can be explainded that the services that prevent terrorism is doing a good job and the relative number of incidents are not that high to prevent tourists from visiting a country.

Taxes on goods and services has no effect on tourism flows. This shows that at least in the western countries the strain of taxes on the service sector is beeing compensated by other money flows and tourist do not feel a lack of coffee or souvenir shops and other small businesses.

\newpage

\section{Structure models}

In this section we will try to obtain more info from the data. We will try to create a model:

\[
\Delta log(A)_{ikt}  = \alpha_{it} + \beta X_{it}^{d} + \gamma X_{kt}^{a} + \epsilon_{it} 
\]

Here 

$A_{ikt}$ - total number of arrivals to country i from country k at time t.

$X_{it}^{d}$ - economic state of country i at time period t (GDP per capita); d stands for domestic

$X_{kt}^{a}$ - economic state of country k at time period t (GDP per capita); a stands for abroad

We have data spaning from 2008 to 2014 that contains information regarding tourism flows from several places around the world to the western countries. For example the flows of tourists to United Kingdom in 2014 were:


```{r, echo=F}
kable(struct[struct$Country=="United Kingdom" & struct$Date=="2014", ], format = "latex", row.names = F)
```

We will save the p values and the coefficients near the explanatory variables and analyze the medians and the distributions of them. 

```{r structure models, include=FALSE, echo=F}

sdt <- struct
sdt <- sdt[!(sdt$Indicator %in% c("Other collective establishments", "Hotels and similar establishments", "Overnight visitors (tourists)", "Total international arrivals", "Same-day visitors (excursionists)", "Nights in all types of accommodation", 
                                  "Specialised establishments", "Private accommodation")), ]

sdt <- sdt[sdt$Country %in% unique(pdata$Country) & sdt$Indicator %in% unique(gdp$Country), ] %>% rename(c("Value" = "Arrivals", "Indicator" = "Origin"))

result.frame <- data.frame()

ddply(sdt, ~Country + Origin, function(xframe){
  
  xframe <<- xframe
  if(dim(xframe)[1] >= 6){
    ## Adding variable for domestic economy
    
    xframe <- merge(xframe, gdp) %>% rename(c("value" = "gdp.d"))
    xframe$Indicator <- NULL
    
    ## Adding variable for foreign economy
    
    gdp.a <- gdp %>% filter(Country==xframe$Origin[1], Date %in% 2008:2014) %>% rename(c("value" = "gdp.a"))
    xframe$gdp.a <- gdp.a[, "gdp.a"]
    
    xframe$gdp.d <- c(NA, diff(log(xframe$gdp.d)))
    xframe$gdp.a <- c(NA, diff(log(xframe$gdp.a)))
    xframe$Arrivals <- c(NA, diff(log(xframe$Arrivals))) 
    
    model <- lm(Arrivals ~ gdp.d + gdp.a, data=xframe)
    
    rez <- data.frame("gdp.a" = model$coefficients[3],
                      "gdp.d" = model$coefficients[2],
                      'p.a' = summary(model)$coefficients[, 4]["gdp.a"], 
                      'p.d' = summary(model)$coefficients[, 4]["gdp.d"])
    
    result.frame <<- rbind(result.frame, rez)
  }
  
})
```

```{r histogram of structure models, echo=F, fig.cap="Histograms", fig.height=8, fig.width=7,  fig.pos="H"}
  
par(mfrow=c(2, 2))

hist(result.frame$gdp.a, xlab="GDP abroad", main="Coefficient near GDP abroad", col="dodgerblue1")
hist(result.frame$gdp.d, xlab="GDP domestic", main="Coefficient near GDP domestic", col="dodgerblue1")

hist(result.frame$p.a, xlab="p - value", main="p - value of GDP abroad", col="dodgerblue1", breaks = 10)
hist(result.frame$p.d, xlab="p - value", main="p - value of GDP domestic", col="dodgerblue1", breaks = 10)

```

All in all we have created 57 structure models. In figure 10 we can see the histograms of the coefficients and the p - values of the coefficients. The median of the coefficient near the GDP per capita abroad variable is 0.329 and domestic is 0.123. This would imply that a bigger impact to the change of tourist arrivals is due to the positive change in abroad economies. The number of times the p value for the abroad coefficient was smaller than 0.2 was 22 and the p value for domestic economy was 14. The main shortcoming of this analysis is that we have very few data points to work with - after differencing we have a time span of 2009 to 2014. 

\newpage

\subsection{Full matrix linear model}

In this section we will be creating a data frames Y and X, where:

$Y_{ikt}$ - the total number of arrivals from country k to country i at time period t.

$X_{itk}$ - both the domestic and abroad economy's GDP per capita, domestic terror attacks, domestic taxes on goods and services.

We will be estimating the $\beta$ in:

$$Y_{itk} = \alpha + \beta X_{itk} + \epsilon$$

We will use the OLS procedure:

$$ \widehat{\beta} = \left( X^{T} X \right)^{-1}X^{T}Y $$

The structure of Y is:

\[Y = \begin{bmatrix}
         y_{1, 1, 2009}    \\
         y_{2, 1, 2009}    \\[0.3em]
         \vdots   \\[0.3em]
         y_{1, 1, 2009}  \\[0.3em]
          \vdots   \\[0.3em]
         y_{1, k, 2014}
  \end{bmatrix} \]
  
The X is a matrix of 6 collumns (intercept collumn and the economic variables) and has the same number of rows as Y.
 
```{r structured panel data, echo=F, include=FALSE}

sdt.sub <- sdt
sdt.sub <- ddply(sdt.sub, ~Country + Origin, function(xframe){
  
  xframe <<- xframe
  if(dim(xframe)[1] >= 6){
    
    ## Adding variable for domestic economy
    
    xframe <- merge(xframe, gdp) %>% rename(c("value" = "gdp.d"))
    xframe$Indicator <- NULL
    
    ## Adding variable for foreign economy
    
    gdp.a <- gdp %>% filter(Country==xframe$Origin[1], Date %in% 2008:2014) %>% rename(c("value" = "gdp.a"))
    xframe$gdp.a <- gdp.a[, "gdp.a"]
    
    ## Adding hours worked 

    h <- hours %>% filter(Country==xframe$Country[1], Date %in% 2008:2014) %>% rename(c("Value" = "hours.worked"))
    xframe$hours <- h[, "hours.worked"]

    ## Adding terror attacks
    
    ter <- terror %>% filter(Country==xframe$Country[1], Date %in% 2008:2014)
    xframe$terror <- ter[, "Terror.attacks"]
    
    ## Adding taxes
    
    taxes <- tax %>% filter(Country==xframe$Country[1], Date %in% 2008:2014) %>% rename(c("Value" = "taxes"))
    xframe$taxes <- taxes[, "taxes"]
    
    xframe$hours <- c(NA, diff(log(xframe$hours)))
    xframe$gdp.d <- c(NA, diff(log(xframe$gdp.d)))
    xframe$gdp.a <- c(NA, diff(log(xframe$gdp.a)))
    xframe$Arrivals <- c(NA, diff(log(xframe$Arrivals))) 
    xframe$terror <- c(NA, diff(xframe$terror)) 
    xframe$taxes  <- c(NA, diff(xframe$taxes))
    
  }
  
    return(xframe)
})

sdt.sub$Origin <- NULL
sdt.sub <- sdt.sub[complete.cases(sdt.sub), ]

model.ols <-  lm(Arrivals ~  gdp.a + gdp.d + terror + hours + taxes, data=sdt.sub)

```

As we can see from table 9, the only significant variables are GDP per capita abroad and hours worked domestically.

```{r, results='asis', echo=F, warning=F}

tab <- xtable(summary(model.ols)$coef, digits=c(3, 3, 3, 3, 3), caption = "OLS model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)
```

```{r, results='asis', echo=F, warning=F}

model.ols <-  lm(Arrivals ~  gdp.a +  hours, data=sdt.sub)
tab <- xtable(summary(model.ols)$coef, digits=c(3, 3, 3, 3, 3), caption = "Final OLS model")
print(tab, type="latex", floating = T, only.contents = F, comment = F)

```

As in the case of panel models, the terror attack variable is not significant. What is more interesting that the number of tourist arrivals does not depend on the local economy but rather on the economic states of countries from which there are visitors. A one percent change in GDP of a country from which tourists are arriving leads to an increase of 0.352 percent in positive change to tourist arrivals.

\newpage

\subsection{Residual analysis}

```{r distribution of residuals of final model, echo=F, fig.cap="Distribution of the residuals", fig.height=7, fig.width=8,  fig.pos="h"}
  
par(mfrow=c(2,1))

res <- residuals(model.ols)

hist(res, col = "cornsilk3", xlab = "Residuals of the structure model", main="Density of the residuals")

qqnorm(res, ylab='Residuals', main = "Q - Q plot for structure model", col="cornflowerblue")
grid()
qqline(res)

```

As we can see from figure 11 we can conclude that the residuals from the structure model graphically look the most normal out of all the models in this analysis so far.

```{r cooks distance, echo=F, fig.cap="Distribution of the residuals", fig.height=7, fig.width=7,  fig.pos="h"}

par(mfrow=c(2, 1))  
plot(model.ols, 4, col="salmon1", lwd=2)
abline(h=6/nrow(sdt.sub), col="dodgerblue1", lwd=2)
legendary2(col="dodgerblue1", "4/number of obs.")
grid()

c.dist <- cooks.distance(model.ols)
r <- stdres(model.ols)
plot(r, type="l", xlab="number of obs.", ylab="Residaul", main="Plot of standartized residuals", col="salmon1")
abline(h=3*sd(r), col="dodgerblue1", lwd=2)
abline(h=-3*sd(r), col="dodgerblue1", lwd=2)
legendary2(col="dodgerblue1", "3 sigmas")
grid()

```

We can clearly see from figure 12 that there are some observations which cook's distance is much larger than the generally accepted rule of 4 divided by the number of observations. We will eliminate the points from our data set who's residual is larger in absolute value than 2.5.


```{r model without influantial points, results='asis', echo=F, warning=F}
  
sdt.sub$res <- abs(r)
sdt.sub <- sdt.sub[sdt.sub$res<2.5, ]
model.ols <-  lm(Arrivals ~  gdp.a +  hours , data=sdt.sub)  
tab <- xtable(summary(model.ols)$coef, digits=c(3, 3, 3, 3, 3), caption = "OLS model without the influential points")
print(tab, type="latex", floating = T, only.contents = F, comment = F)

```

Eliminating the influential points and re-regressed the model we got that the variable hours is not significant after all. Also we can see that from figure 13 that the residuals are visually normal. The Shapiro-Wilk test's statistic is 0.039 which with alpha level of 0.01 would imply normality.

```{r distribution of residuals of final model 2, echo=F, fig.cap="Distribution of the residuals without influential points", fig.height=7, fig.width=7,  fig.pos="h"}

model.ols <-  lm(Arrivals ~  gdp.a, data=sdt.sub)    
par(mfrow=c(2,1))

res <- residuals(model.ols)

hist(res, col = "cornsilk3", xlab = "Residuals of the structure model", main="Density of the residuals")

qqnorm(res, ylab='Residuals', main = "Q - Q plot for structure model", col="cornflowerblue")
grid()
qqline(res)

```

```{r, results='asis', echo=F, warning=F}

model.ols <-  lm(Arrivals ~  gdp.a, data=sdt.sub)
tab <- xtable(summary(model.ols)$coef, digits=c(3, 3, 3, 3, 3), caption = "Final OLS model without influential points")
print(tab, type="latex", floating = T, only.contents = F, comment = F)

```

The final structure model is in table 12.


\newpage

\section{Conclusion}

After testing various models we can make several conclusions regarding tourism models for the countries in this analysis:

$\bullet$ Random effects, fixed effects, dynamic panel models and structure models showed that terror attacks did not have an influential impact to overall change to tourist arrivals.

$\bullet$ Increasing GDP per capita in a given country does not influence tourist arrivals to that country.

$\bullet$ Percentage change of taxes on goods and services in country i does not influence the rate of change in tourist arrivals.

$\bullet$ The most significant variable regarding the percentage change in tourist arrivals to country i is the percentage change in GDP per capita in abroad countries from whom the tourists come to the country i.

$\bullet$ According to our model, every country is better off in the sense of positive tourist arrivals change when the economies are growing.

\newpage

\section{Tests}

In this paper there were implemented two tests in order to inspect the panel models: the F test for individual effects and the Haussman test to determine which model - random or fixed effects - is more suitable. 

\subsection{F test}

The f test [see @F.test, pp.3-4] that is used with the panel models in R statistical package tests the following hypothesis: assume that we have two models: 

restricted
$$y_{it} = \alpha + \beta X_{it} + \epsilon_{it}$$
and unrestricted:

$$y_{it} = \alpha + u_{i} + \beta X_{it} + \epsilon_{it}$$

Then the hypothesis is 


\[\begin{array}{lr}
        H_{0}: u_{i} =0,  \forall i \\
        H_{1}: \exists i:    u_{i} \neq 0,  
        \end{array}\]

The alternative hypothesis would mean that there exists a country where adding the term $u_{i}$ the r-squared statistic increases. 

To test this hypothes we use the F statistic (hence the name of the test):

$$F = \dfrac{(ESS_{R} - ESS_{U})/(N- 1) }{ESS_{U}/((T-1)N- K)}$$

Here 

$ESS_{U}$ - error sum of squares of the unrestricted model

$ESS_{R}$ - error sum of squares of the restricted model

$N$ - number of countries

$T$ - number of time periods

Under the assumption that the residuals are gaussian then the F statistic is distributed by the $\textbf{F}$ distribution with N-1, N(T-1) - K degrees of freedom.

\newpage

\subsection{Hausman test for panel models}

The Hausman test [see @Hausman.test, pp.39-40] is used to determine whether $X_{it}$ are uncorrelated with $\alpha_{i}$. To test this we denote two estimators:

$\widehat{\beta_{FE}}$ - fixed effects estimate. 

$\widehat{\beta_{RE}}$ - random effects estimate.

The hypothesis is then:

 Estimator                $H_{0}$ holds              $H_{0}$ is rejected  
-------                  -------------              -----------------
$\widehat{\beta_{RE}}$    consistent;                 Inconsistent
                          efficient
$\widehat{\beta_{FE}}$    consistent;                  consistent       
                          inefficient

-------                  -------------              -----------------


Since under null hypothesis $\widehat{\beta_{RE}}$ is efficient then 

$$Var(\widehat{\beta_{FE}} - \widehat{\beta_{RE}}) =Var(\widehat{\beta_{FE}}) - Var(\widehat{\beta_{RE}})$$
Then the Hausman statistic is defined as:

$$H = \left( \widehat{\beta_{FE}} - \widehat{\beta_{RE}} \right) ^{T} \left(Var(\widehat{\beta_{FE}}) - Var(\widehat{\beta_{RE}})\right) \left( \widehat{\beta_{FE}} - \widehat{\beta_{RE}} \right)$$

and 

$$H \overset{asy}{\sim} \chi^2$$

The main idea of this test is that if $Var(\widehat{\beta_{FE}})$ is smaller than  $Var(\widehat{\beta_{RE}})$ it means that the H statistic will be "large" and the null hypothesis will be rejected. This in turn means that $\widehat{\beta_{RE}}$ is no efficient because there exists another statistic with a smaller variance.


\newpage

# References



