---
output:
 pdf_document: 
    fig_caption: yes
    keep_tex: yes
    number_sections: no
    # latex_engine: xelatex
   # header_includes: \usepackage{float}

---

```{r setup, include=FALSE}
library(plm)
library(RCurl)
library(dplyr)
library(plyr)
library(emibase)
library(plotrix)
library(mgcv)
library(reshape2)
library(readxl)
library(knitr)
library(rworldmap)
library(cshapes)
library(countrycode)
library(texreg)
library(ggplot2)
library(directlabels)
library(splines)
library(MASS)
source('functions.R')
```

\vskip 20pt
\centerline{\bf \large VILNIUS UNIVERSITY}
\bigskip
\centerline{\large \textbf{FACULTY OF MATHEMATICS AND INFORMATICS}}
\vskip 120pt
\centerline{\bf \Large \textbf{BACHELOR THESIS}}
\vskip 50pt
\begin{center}
{\bf \LARGE Tourism model in OECD countries}

\vspace{4mm}

\end{center}
\vskip 120pt
\centerline{\Large Eligijus Bujokas}
\vskip 120pt
\centerline{\large \textbf{VILNIUS (2016.12.07)}}

\newpage

\tableofcontents

\newpage

```{r constants, include=FALSE}
years.to.survey <- 1970:2015
path <- "data/"
# regions <- read.csv(file="input/regions.csv", stringsAsFactors = F)
createdir("plots")
# createdir("input")
createdir("data")
createdir("output")
devtools::install_url("http://cran.r-project.org/src/contrib/rprojroot_1.2.tar.gz") ## MAY RESOLVE TROUBLESHOOTS
knitr::opts_chunk$set(fig.pos = 'h')
```

```{r data download and reading it, include=FALSE, echo=F}
# we will save this data to the 'data' folder
download.terror(years.to.survey, "data/") 
dt <- read.terror("data/terror/")
dt <- arrange(dt, Date)
na.index <- which(apply(dt, 1, function(x){ all(is.na(x)) }))
if(length(na.index)!=0) dt <- dt[-na.index, ]
write.csv(dt, file="output/terror raw data.csv", na="", row.names=F)

## Tourism data

dt.tour <- download.tourism("data/")

## Economic data

dt.eco <- download.economy("data/")
dt.eco <- make.economy.great.again(dt.eco)

## OECD data on hours worked

dt.oecd <- get_dataset("ANHRS")
decode  <- get_decoder("data/")  

# Decoding country names

dt.oecd <- ddply(dt.oecd, ~COUNTRY + obsTime + EMPSTAT, function(xframe){
  
  xframe <<- xframe
  real.name <-  decode[decode$CODE==xframe$COUNTRY[1], "Country"]
  if(length(real.name)==0) real.name <- xframe$COUNTRY[1]
  xframe$COUNTRY <- real.name
  return(xframe)
  
})

dt.oecd <- dt.oecd[dt.oecd$EMPSTAT=="TE", ]
dt.oecd[dt.oecd$COUNTRY=="KOR", "COUNTRY"] <- "South Korea"
# dt.oecd[dt.oecd$COUNTRY=="Slovak Republic", "COUNTRY"] <- "Slovakia"


```

```{r descriptive statistics, include=FALSE, echo=F, warning=F}
# # we will save this data to the 'data folder'
# summary(dt$Fatalities) ## As we can see there is a huge incident involving 1382 deaths
# summary(dt$Injured)    ## The mean of injured people is higher than those that were killed. The biggest incident is involving 4000 people
# dt[which(dt$Fatalities==max(dt$Fatalities, na.rm=T)), ] # The deadliest attack was the attack in Nepal in 2004
# dt[which(dt$Injured==max(dt$Injured, na.rm=T)), ]       # The most injuries was caused by the attack in Kenya in 1998
# 
# sum(dt$Injured, na.rm=T) 
# sum(dt$Fatalities, na.rm=T) 
# 
# 
# ## Creating a pie chart
# 
# pdf(file = "plots/Injuries and deaths pie chart.pdf", width = 15, height = 10)
# 
# pie.chart(dt$Fatalities, c(0, 1, 5, 100), cols=c("dodgerblue1", "orchid4", "gray48", "peru"), 
#           title = "Fatalities")
# 
# pie.chart(dt$Injured, c(0, 1, 5, 100), cols=c("dodgerblue1", "orchid4", "gray48", "peru"), 
#           title = "Injuries")
# dev.off()
# 
# ## Overall trend
# 
# pdf(file = "plots/daily monthly yearly casualties.pdf", width = 15, height = 10)
# 
# # excluding daily because of long calculation times
# 
# for(fr in c("monthly", "yearly")){
#   data.to.plot <- aggregate.terror(dt, freq = fr)
#   
#   points <- cbind(data.to.plot$Injured, data.to.plot$Killed)
#   
#   time.span <- seq(years.to.survey[1],
#                    years.to.survey[length(years.to.survey)],
#                    length.out = dim(data.to.plot)[1])
#   
#   grid.frame(x = time.span, y = points, grid.lty = 2, grid.col = "royalblue")
#   
#   matplot(x= time.span, y=points, add=T, type = "o", pch=20, lwd = 2, lty = 1, cex=1, 
#               col=c("firebrick3", "dodgerblue3"))
#   legendary2(c("Injured", "Killed"), col=c("firebrick3", "dodgerblue3"))
#   mtext(paste("Frequency is", fr), line=1, adj = 0, font = 2, cex = 1.25, col="darkslateblue")
# }
# dev.off()
# 
# ## Graph by country
# 
# countries <- dt$Country %>% unique()
# 
# all.countries <- data.frame()
# 
# for(cn in countries){
#   sub.set <- dt[dt$Country==cn, c("Country", "Injured", "Fatalities")]
#   x <- c(cn,
#    sum(sub.set$Injured, na.rm=T),
#    sum(sub.set$Fatalities, na.rm=T)) %>% as.data.frame() %>% t()
#   x[, 2:3] <- as.numeric(x[, 2:3])
#   all.countries <- rbind(all.countries, x)
# }
# 
# colnames(all.countries) <- c("Country", "Injuries", "Fatalities")
# 
# # Plottting according to regions
# 
# pdf(file = "plots/Injuries deaths by region.pdf", width = 20, height = 10)
# for(reg in unique(regions$Region)){
#   for(type in c("Injuries", "Fatalities")){
#     countries <- intersect(regions[regions$Region==reg, "Country"], unique(all.countries$Country))
#     data.set  <- all.countries[all.countries$Country %in% countries, c("Country", type)]
#     data.set$Country <- as.character(data.set$Country)
#     data.set  <- arrange(data.set, Country)
#     codeID <- regions[regions$Region==reg & regions$Country %in% as.character(data.set$Country), ]
#     codeID <- arrange(codeID, Country) 
#     values <- data.set[, 2] %>% as.character() %>% as.numeric()
#     names(values) <- codeID$CountryCode
#     values <- sort(values)
#     
#     barplot(values, col='deepskyblue3',  horiz=T , las=1, main = paste(type, "in", reg))
#   }
# }
# dev.off()

```

```{r data for panel model, include=FALSE, echo=F}

# In this section we will be creating a data frame for a VAR model or a panel model

master.data <- aggregate.terror.by.country(dt)

dt.tour <- rename(dt.tour, c("Year" = "Date", "CountryName" = "Country"))
dt.eco <- rename(dt.eco, c("Year" = "Date", "CountryName" = "Country"))

dt.tour$Country <- gsub("Korea, Rep.", "South Korea", dt.tour$Country)
dt.eco$Country <- gsub("Korea, Rep.", "South Korea", dt.eco$Country)

master.data <- merge(master.data, dt.tour)

for(prod in unique(dt.eco$Indicator)){

  tmp <- dt.eco[dt.eco$Indicator==prod, c("Country", "Date", "value")]
  tmp <- rename(tmp, c("value" = prod))
  master.data <- merge(master.data, tmp)

}

# Merging with OECD data

dt.oecd <- plyr::rename(dt.oecd, c("COUNTRY" = "Country", "obsTime" = "Date", "obsValue" = "Working hours"))
dt.oecd <- dt.oecd[, c("Country", "Date", "Working hours")]

master.data <- merge(master.data, dt.oecd)

# Dropping incomplete rows

master.data <- master.data[complete.cases(master.data), ]

master.data <- arrange(master.data, Country)

# Adding number of neighbours

neighbours <- find.neighbours(unique(master.data$Country), path=path)

master.data <- merge(master.data, neighbours, sort=F)
master.data$Numb.neighbour <- as.numeric(as.character(master.data$Numb.neighbour))


master.data <- plyr::rename(master.data, c('GDP per capita (current US$)' = "GDP.per.capita"))
master.data <- plyr::rename(master.data, c('Working hours' = "Hours.worked"))

```

\section{Countries}

The countries which will be analyzed are all members of the OECD (\textbf{Organisation for Economic Co-operation and Development}). Currently there are 35 member states but in this analysis we will be adding Lithuania and Costa Rica because it is strongly believed that these countries will soon join OECD.

```{r map of OECD countries, echo=F, fig.width=8, fig.height=8, fig.cap= "Map of the OECD countries", echo=F, warning=F, include=T}

OECD.cn <- data.frame(country = unique(master.data$Country), 
                      OECD = rep.int("OECD countries", 
                                     length(unique(master.data$Country))))
malMap <- joinCountryData2Map(OECD.cn, joinCode = "NAME",
  nameJoinColumn = "country", verbose = F, suggestForFailedCodes = F)

mapCountryData(malMap, nameColumnToPlot="OECD", catMethod = "categorical",
  missingCountryCol = gray(.8), oceanCol = 'cyan', mapTitle = "Political map of the world")


```

\newpage

\subsection{Total number of arrivals}

To have a better picture of the tourist flows to the OECD countries we can visualize the data.

```{r arrivals visualization, echo=F, fig.cap="Total number of tourist arrivals", fig.height=8, fig.width=11, fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "Total.Arrivals"), ]

dt.a$Date <- as.numeric(dt.a$Date)
dt.a$Total.Arrivals <- dt.a$Total.Arrivals/1000

p <- ggplot(dt.a, aes(x=Date, y=Total.Arrivals, group=Country, colour=Country)) + geom_line(size=1.2) + ylab("Total Arrivals, thousands") + theme_bw()  + theme(legend.position="none") + scale_x_continuous(expand = c(0, 3)) + geom_dl(aes(label = Country), method = list(dl.combine("last.bumpup"), cex = 1)) 

plot(p)

```

To see the overal trend from all of these countries is quite difficult so we will plot the average of tourist arrivals in each year.

```{r arrival trend, echo=F, fig.cap="Trend of tourists arrivals", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "Total.Arrivals"), ]

dt.a <- ddply(dt.a, ~Date, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country='OECD', Date=xframe$Date[1], Total.Arrivals=mean(xframe$Total.Arrivals, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a$Total.Arrivals <- dt.a$Total.Arrivals/1000
dt.a$Date <- as.numeric(dt.a$Date)

p <- ggplot(dt.a, aes(x=Date, y=Total.Arrivals, group=Country, colour=Country, label=Country)) + geom_line(size=1.2) + ylab("Total Arrivals, thousands") + theme_bw() + scale_x_continuous(expand = c(0, 4)) + geom_dl(aes(label = Country), method = list(dl.combine("first.polygons"), cex = 1)) + geom_dl(aes(label = Country), method = list(dl.combine("last.polygons"), cex = 1)) + stat_smooth(se = FALSE, method = "lm", formula = y ~ ns(x,3),aes( colour="Square trend"), linetype="dashed") +
scale_colour_manual(values = c("coral3", "darkblue"),
                       guide = guide_legend(override.aes = list(
                         linetype = c("solid", "dashed"),
                         shape = c(16, 16)), title="Legend")) 

plot(p)

```

From figure 3 we can clearly see that the overal mean through the years is increasing rapidly. In this paper we will try to model these arrival values and identify the biggest factors of this process. 

\newpage
\section{Modelling}

The aim of this paper is to create a model to help explain and forecast the total number of arrivals of tourists to a given OECD country. We want to find the function \textbf{f} in the equation:

\[
log(A)_{it}  = \textbf{f}(log(GDP)_{it}, terror_{it}, log(hours)_{it}, neighbours_{i})
\]

Here 

$A_{it}$ - total number of arrivals to a country i at time period t, thousands

$GDP_{it}$ - GDP per capita at a country i at time period t, thousands 

$terror_{it}$ - total number of terrorist attacks at country i at time period t

$hours_{it}$ - average numbers of hours worked annualy at country i at time period t

$neighbours_{i}$ - a factor variable indicating whether a country has a "good" geographical location or not. The measure of goodness is whether a country has at least 5 or more surounding countries which the smallest distance to country i is smaller than 100 km.


\subsection{Motivation behind variable selection}

\subsubsection{Gross domestic product per capita}

The variable $GDP$ was chosen to reflect the economic situation and the standart of living in a given country. GDP per capita is calculated by taking the  total output of a country and dividing it by the number of people in the country. The division by the number of people helps us to compare economies between countries. 

```{r OECD GDP per capita, echo=F, fig.cap="GDP per capita of OECD", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "GDP.per.capita")]

dt.a <- ddply(dt.a, ~Date, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country='OECD', Date=xframe$Date[1], GDP.per.capita=mean(xframe$GDP.per.capita, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a <- rename(dt.a, c("GDP.per.capita" = "value"))
dt.a$Date <- as.numeric(dt.a$Date)

p <- ggplot(dt.a, aes(x=Date, y=value, group=Country, colour=Country, label=Country)) + geom_line(size=1.2) + ylab("GDP per capita, USD") + theme_bw() + scale_x_continuous(expand = c(0, 4)) + geom_dl(aes(label = Country), method = list(dl.combine("first.polygons"), cex = 1)) + geom_dl(aes(label = Country), method = list(dl.combine("last.polygons"), cex = 1))  +
scale_colour_manual(values = c("coral3"),
                       guide = guide_legend(override.aes = list(
                         linetype = c("solid"),
                         shape = c(16)), title="Legend")) 

plot(p)

```

\newpage

\subsubsection{Number of terrorist attacks}

In recent years there has been an increase in terrorist attacks in the OECD member countries: France, Germany, Turkey and others have suffered attacks, especially from the group called ISIS. It is quite straightforward to think that in light of these attacks tourists will be trying to avoid countries with a higher chance of a terrorist attack. We will want to know whether terrorist attacks is a significant variable to tourist arrivals and how much of an impact it has. 

```{r visualizing terror, echo=F, fig.cap="Terrorism in OECD countries", fig.height=6, fig.width=7,  fig.pos="H"}

dt.a <- master.data[, c("Country", "Date", "Terror.attacks")]

dt.a <- ddply(dt.a, ~Country, function(xframe){
  
  xframe <<- xframe
  
  results <- data.frame(Country=xframe$Country, Date="Total", Terror.attacks=sum(xframe$Terror.attacks, na.rm=T))
  xframe <- rbind(xframe, results)
  return(xframe[nrow(xframe), ])

})

dt.a <- t(dt.a[, c("Country", "Terror.attacks")])
# dt.a[2, ] <- as.numeric(dt.a[2, ])
nn <- dt.a[1, ]
dt.a <- dt.a[-1, ] %>% as.data.frame() %>% t()
dt.a <- apply(dt.a, c(1,2), function(x) {as.numeric(x)})
names(dt.a) <- nn
dt.a <- sort(dt.a)
par(mai=c(1,2,1,1))
barplot(dt.a , col='deepskyblue3',  horiz=T , las=1, cex.names = 0.7, main="Total number of attacks")

```

\newpage

\subsection{Panel data model}

Panel data provides information on individual behaviour, both across individuals and over time - the data has cross-sectional and time-series dimensions. Panel data includes N individuals over T regular time periods. We can denote the model simmilary like a typical bivariate reggression model:

\[
  y_{it} = \alpha_{i} + \beta X_{it} + u_{it}
\]

The difference from a typical reggression model is that there are the $i$ and the $t$ terms, where $i$ denotes the individual and $t$ denotes the time period. The construction of a panel model starts with appropriately 'stacking' the data set. Assuming that y is the dependant variable and X is the matrix of the explanatory variables we need to have the following structure: 

\[
  Y = y_{i} = \begin{bmatrix}
         y_{1}    \\
         y_{2}    \\[0.3em]
         \vdots   \\[0.3em]
         y_{N}     
  \end{bmatrix} = \begin{bmatrix}
         e    \\
         0    \\[0.3em]
         \vdots   \\[0.3em]
         0     
  \end{bmatrix}  \alpha_{1} + \begin{bmatrix}
       0    \\
       e    \\[0.3em]
       \vdots   \\[0.3em]
       0    
\end{bmatrix}  \alpha_{2} + \cdots + \begin{bmatrix}
       0    \\
       0    \\[0.3em]
       \vdots   \\[0.3em]
       e     \\[0.3em]
\end{bmatrix}  \alpha_{N} + \begin{bmatrix}
       x_{1}    \\[0.3em]
       x_{2}    \\[0.3em]
       \vdots   \\[0.3em]
       x_{n}     \\[0.3em]
\end{bmatrix} \beta +   \begin{bmatrix}
       u_{1}    \\[0.3em]
       u_{2}    \\[0.3em]
       \vdots   \\[0.3em]
       u_{n}     \\[0.3em]
\end{bmatrix}
\]

Here

\[
y_{i} = \begin{bmatrix}
       y_{i1}    \\
       y_{i2}    \\[0.3em]
       \vdots   \\[0.3em]
       y_{iT}     \\[0.3em]
\end{bmatrix} x_{i} = \begin{bmatrix}
       x_{1i1} & x_{2i1} & \cdots & x_{Ki1}    \\
       x_{1i2} & x_{2i2} & \cdots & x_{Ki2}   \\[0.3em]
       \vdots & \vdots & \cdots & \vdots   \\[0.3em]
      x_{1iT} & x_{2iT} & \cdots & x_{KiT}      \\[0.3em]
\end{bmatrix}
\]

For example, the 'head' and the 'tail' of the raw data which will me analysed looks like this:

```{r visualising data table, echo=F}

pdata <- plm.data(master.data, index = c("Country", "Date")) 

pdata$Total.Arrivals <- log(pdata$Total.Arrivals)
pdata$GDP.per.capita <- log(pdata$GDP.per.capita)
pdata$Hours.worked   <- log(pdata$Hours.worked)

## Creating a factor indicating a "good" geographical location

pdata$Numb.neighbour <- sapply(pdata$Numb.neighbour, function(x){
  
  if(x < 5) x <- 0
  if(x >=5) x <- 1
  return(x)
  
})

pdata$Numb.neighbour <- as.factor(pdata$Numb.neighbour)
kable(head(pdata), format = "latex", row.names = F)
kable(tail(pdata), format = "latex", row.names = F)

```


```{r pooling panel model, include=FALSE, echo=F}

model <- plm(formula = Total.Arrivals ~ Terror.attacks + GDP.per.capita + Hours.worked + Numb.neighbour , data=pdata, model="pooling")
# summary(model)
```

\subsubsection{Pooling estimator}

The first model which we will be examining will be the simple OLS model which ignores the panel data structure. The equation is then:

\[
y = \alpha + X \beta + \epsilon
\]

And the estimator is derived solving the equation:

\[
\widehat{\beta} = \left( X X^{T} \right)^{-1}X^{T}y
\]


```{r, results='asis', echo=F, warning=F, fig.pos="h"}
  texreg(model, caption = "OLS model", custom.coef.names = c("Intercept", "log attacks", "log GDP", "log Hours", "neighbours"))
```

According to this model, there is a positive influence to tourism caused by terrorist attacks, which is not logical. Given that this model does not take into consideration the panel structure of the data, we will not be using straighforward OLS in our analysis. 

\subsubsection{Random effects model}

In this section we will be creating a random effects panel data model. Recall equation (1). The $\alpha_{i}$ is called the \textit{individual effect}.
If $\alpha_{i}$ does not correlate with $X_{it}$ then the panel model can be reffered to as a random effects model. 












