---
output:
  pdf_document:
    fig_caption: yes
    number_sections: no
   # toc: true
#bibliography: library.bib
---

```{r setup, include=FALSE}
library(dplyr)
library(plyr)
library(emibase)
library(plotrix)
source('functions.R')
```

\vskip 20pt
\centerline{\bf \large VILNIUS UNIVERSITY}
\bigskip
\centerline{\large \textbf{FACULTY OF MATHEMATICS AND INFORMATICS}}
\vskip 120pt
\centerline{\bf \Large \textbf{BACHELOR THESIS}}
\vskip 50pt
\begin{center}
{\bf \LARGE Global terrorism and the economy}

\vspace{4mm}

\end{center}
\vskip 120pt
\centerline{\Large Eligijus Bujokas}
\vskip 120pt
\centerline{\large \textbf{VILNIUS (2016.12.07)}}

\newpage




\tableofcontents

\newpage

```{r constants, include=FALSE}
years.to.survey <- 1970:2015
```

```{r data download and reading it, include=FALSE, echo=F}
# we will save this data to the 'data folder'
download.terror(years.to.survey, "data/") 
dt <- read.terror("data/")
```

```{r descriptive statistics, include=FALSE, echo=F, warning=F}
# we will save this data to the 'data folder'
summary(dt$Fatalities) ## As we can see there is a huge incident involving 1382 deaths
summary(dt$Injured)    ## The mean of injured people is higher than those that were killed. The biggest incident is involving 4000 people
dt[which(dt$Fatalities==max(dt$Fatalities, na.rm=T)), ] # The deadliest attack was the infamous 9/11 attack
dt[which(dt$Injured==max(dt$Injured, na.rm=T)), ]       # The most injuries was caused by the attack in Kenya in 1998

sum(dt$Injured, na.rm=T) # The total people injured between 1970 and 2015 worldwide is 125718
sum(dt$Fatalities, na.rm=T) # The total people killed between 1970 and 2015 worldwide is 97233


## Creating a pie chart

pdf(file = "plots/Injuries and deaths pie chart.pdf", width = 15, height = 10)
par(mfrow=c(2,1))

pie.chart(dt$Fatalities, c(0, 1, 5, 100), cols=c("dodgerblue1", "darkslategray2", "gray48", "khaki2"), 
          title = "Fatalities")

pie.chart(dt$Injured, c(0, 1, 5, 100), cols=c("dodgerblue1", "darkslategray2", "gray48", "khaki2"), 
          title = "Injuries")
dev.off()
```





